Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-16 latex/main.tex
edits from Shing

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

Before characterizing recombination events in detail,
we sought to remove probable artifacts.
Systematic errors in SARS-CoV-2 sequences
associated with sequencing protocols \cite{Turakhia2020}
can complicate recombination detection \cite{DeMaio2024}.
% Explain to readers what convincing genomic evidence looks like.
By inspecting copying patterns at sites that differ between recombinant parents,
we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.
Artifactual recombination events can be characterized
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
When counting such loci, we considered nearly adjacent sites
(separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci,
which could lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
The copying patterns depicted Figure~\ref{fig:copying_pattern_examples}
illustrate the genomic evidence for three recombination events identified as artifactual,
compared to a marginally robust one.

Evidence that our simple filter effectively flags artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events that we
identify as artifactual. This primer scheme was used to sequence XX of the
samples associated with an artifactual recombination event, in contrast to 15
of the ``robust'' events (Figure 2A), strongly implicating it as a common source
of problematic data.

>>>>

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
These comparisons focus on the Pango X lineages associated with
the Type 1 recombination events (Table~\ref{tab:pango_x_lineages})
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies \cite{Alfonsi2024,Li2024CovRecomb}.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 16 (87.5) & 14 / 16 (87.5) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}

@@

Before characterizing recombination events in detail,
we sought to remove probable artifacts.
Systematic errors in SARS-CoV-2 sequences
associated with sequencing protocols \cite{Turakhia2020}
can complicate recombination detection \cite{DeMaio2024}.
% Explain to readers what convincing genomic evidence looks like.
By inspecting copying patterns at sites that differ between recombinant parents,
we find a convincing signal when treating adjacent and near-adjacent sites as a single locus.

Three copying patterns in Figure~\ref{fig:copying_pattern_examples}
illustrate the evidence for three recombination events identified as artifactual.
Explain the three bad patterns:

Compared to a marginally robust one.

Artifactual recombination events can be characterized
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.

When counting such loci, we considered nearly adjacent sites
(separated by at most three bases) as single loci,
to avoid over-counting support from correlated loci,
which could lead to artificial signals of recombination.
Also, when counting supporting loci for a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 386 ``robust'' events (Figure~\ref{fig:recombinant_qc}).
Evidence that our simple filter effectively flagged artifacts can be seen in the
clear association between the AmpliSeq V1 kit and recombination events
that we identify as artifactual.
This primer scheme was used to sequence XX of the samples
associated with an artifactual recombination event,
in contrast to 15 of the ``robust'' events (Figure~\ref{fig:recombinant_qc}),
strongly implicating it as a common source of problematic data.
